Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-07-22 analysis/paper/paper.qmd
updates

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024]. Early descriptions of spheroids were provided by @clark1955 and @kleindienst1962, and @leakey1971 later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan. Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex. Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].

Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].

At present, the technological category and function of spheroids remain central topics of debate. <!-- BM: excellent, it is good to introduce the debate very explicitly and early in the paper, this is engaging for the reader --> Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979]. Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production [@sahnouni1997]. Another perspective holds that spheroids were naturally produced through high-intensity battering <!-- BM: this is a little unclear, does 'naturally' mean 'an unintentional side-effect of people using them to batter other materials', or 'entirely unrelated to human activity, resulting from natural processes'? can you edit to clarify -->   [@schick1994; @mora2005; @clark1955]. In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy [@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014].

Rather than reflecting a single functional or technological model, these varying interpretations may instead capture regional variability shaped by differences in raw materials, morphologies, and reduction strategies. As @Cabanes2024 noted, spheroids may not represent a homogenous category. This highlights the need to move beyond generalized typological categories and instead engage more directly with local contexts and specific assemblages. Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.

In this study, we demonstrate a novel approach to investigating the technological status of spheroids with a case study from the Middle Pleistocene site of Qianshangying in northern China. Among an assemblage previously classified as a simple core-flake industry, we identified a group spheroids and subspheroids. We observed notable differences between these artifacts and multifacial cores or polyhedrons <!-- BM: not sure what this next phrase means, I recommend you delete or edit to clarify:  "which are stone artifacts through typical flaking strategies" --> and hypothesized distinct reduction sequences. Our goal was to quantitatively evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes that were distinct from the multifacial cores in the assemblage. To accomplish our goal we conducted quantitative morphological and technological analyses of the spheroids recovered from Qianshangying. Our results contribute the debate about whether the concept and manufacture of spheroids represent a distinct technological tradition, separate from conventional flake-based strategies. Our approach includes the first open source and reproducible application of spherical harmonic analysis to an archaeological stone artefact assemblage. These analyses help clarify the technological strategy behind spheroid production at Qianshangying, within the broader context of Middle Pleistocene lithic practices in northern China.

@@

To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying. Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures. Addressing this question is crucial for achieving our goal of determining whether or not spheroids represent a distinct technological category, separate from exhausted cores.

@@

We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

@@

We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree). This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon <!-- BM: please briefly define this --> . After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.

@@

Diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098). Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface. Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form. Phase two was peripheral centripetal exploitation. In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation strategy. A notable feature is the angle between the face's ventral surface and platform (i.e. core edge angle), which often exceeds 90°, forming a secant plan \[citation\] rather than parallel plan. This phase contributes significantly to the increasingly spherical shape of the artefact. Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality. It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three.

<!-- insert QSYA-1098-4.png and QSYA-2797-0.png about here, with the texure paired with the phase annotations figures, like a 2x2 panel, add a legend to indicate the meaning of the colours -->

@@

Multifacial cores and polyhedrons display a wide and dispersed distribution in the ternary plot, without forming any distinctive cluster. Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization and reflect a SSDA. Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence. This distinction highlights the fundamental technological difference between polyhedrons and spheroids at Qianshangying. Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.

@@

According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in Qianshangying. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size). This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.

Therefore, it can be inferred that the Qianshangying hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the Qianshangying facted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].

The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation]. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013]. Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015]. Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product. <!-- BM: this is well written and an appropriate discussion, but I think we should be prepared for some peer reviewers to get excited and try to defend the Acheulean as the very first time shaping strategies are evident! Probably because they have published a paper on that topic and don't want to be proved wrong -->

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production (exhausted cores), or as targeted end products of shaping strategies. In the case of Qianshangying facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd <!-- BM:is the site name correct here? --> and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework [@leakey1971].

Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.

However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores. In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns. At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable. However, it remains unclear whether this typological transformation was intentional or unplanned. In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids. One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.

@@

In China, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids_distribution_China. In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene. In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site in Inner Mongolia [@Wang2010Jinstai]. The widespread occurrence of spheroids across time and space may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China. However, that many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, the Xujiayao site alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities. Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China. In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China. Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun]. For example, at the Dingcun site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz. In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].

Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., Mujiaqiao, [@Wei1984]); pecking, where two blanks are struck together to round off edges—sometimes by splitting a cobble and using the halves against each other (e.g., Dingcun, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., Zhoukoudian, [@NIGPAS1985PekingMan]). Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids]. Many sites, such as Xujiayao and Dingcun, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan]. No clear relationship has yet been established between raw material types and specific production methods.

At Dingcun and Mujiaqiao, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun]. Mujiaqiao specimens retain parts of the cortex, while at Dingcun, flakes with large exterior and small interior platform angles were removed from spheroid blank to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984]. Of the six sites exhibiting similar strategies to QSY spehroids, all but Mujiaqiao are located within the Fen River and Lishui River regions. These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe]. The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy. This is a significant observation, as it implies that spheroid production techniques similar to those at QSY are more frequently associated with lithic assemblages characterized by intentional shaping strategies.

From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites. However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids_distribution_China). Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as Xujiayao, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology. Rather, we argue that faceted spheroids reflect a complex shaping strategy. Their presence within Acheulean contexts is understandable, as spheroids and LCTs share certain core technological features[@leakey1971; @Cabanes2024; @isaac1986foundation]. When faceted spheroids appear in more rudimentary industries or assemblages, it suggests that conceptual templates and advanced technological capacities were already present within those contexts.

This pattern aligns with spheroid findings in other regions of the world. Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan. Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence]. The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019]. Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020]. In particular, faceted spheroids, which reflect a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean—or perhaps even foreshadowing their emergence.

>>>>

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024]. Early descriptions of spheroids were provided by @clark1955 and @kleindienst1962, and @leakey1971 later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan. Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex. Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].

Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another, due to differences in observers’ experience and interpretive criteria. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].Therefore, it is unreasonable to treat them as a coherent group.

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979]. Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production [@sahnouni1997]. Another perspective holds that spheroids were unconsciously produced through high-intensity battering [@schick1994; @mora2005; @clark1955]. In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy [@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014].

At present, the debate remains stalled, primarily due to the lack of clear typological criteria for distinguishing spheroids intentionally shaped according to conceptual templates from those that emerged as by-products of flaking or percussive activities. Regional studies continue to rely heavily on individual researchers’ observational judgment and experiential knowledge. Furthermore, current typological approaches offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites. At the same time, conclusions drawn from their own analyses are rarely subject to external validation or replication.

To address these challenges, we developed a set of quantitative methods combining morphological and technological analyses to investigate spheroid production strategies. While previous morgphlogical measures have focused on the shape of spheroids close to perfect shphere, contributing to discussions of spheroid standardization, they cannot reliably reflect production strategies, as highly spherical forms may result from diverse production goals. We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks. We have creatively modified the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons—this is the first attempt of its kind in the field of archaeology. Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production strategies.

These methods were applied to a case study of spheroids recovered from the previously excavated site of Qianshangying (QSY) of Northern China. In the following part, we assessed whether the QSY spheroids exhibit a patterned production process and standardized morphology, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes distinct from the multifacial cores and polyhedrons within the assemblage. This study provides insights into the production strategy of spheroids at the site. At the same time, we situated the Middle Pleistocene QSY site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China. More importantly, we offer a practical and integrated framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.

@@

To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in QSY. Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures. Addressing this question is crucial for achieving our goal of determining whether or not spheroids represent a distinct technological category, separate from exhausted cores.

@@

We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the QSY assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

@@

We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree). This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts. After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.

@@

Diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098). Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface. Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form. Phase two was peripheral centripetal exploitation. In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation strategy. A notable feature is the angle between the face's ventral surface and platform (i.e. core edge angle), which often exceeds 90°, forming a secant plan rather than parallel plan. This phase contributes significantly to the increasingly spherical shape of the artifact. Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericity. It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three (\@fig-shaping-process).

@@

Multifacial cores and polyhedrons display a wide and dispersed distribution in the ternary plot, without forming any distinctive cluster. Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization. Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence. This distinction highlights the fundamental technological difference between polyhedrons and spheroids at QSY. Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.

@@

According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size). This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.

Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the QSY facted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].

The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation]. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013]. Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015]. Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product. <!-- BM: this is well written and an appropriate discussion, but I think we should be prepared for some peer reviewers to get excited and try to defend the Acheulean as the very first time shaping strategies are evident! Probably because they have published a paper on that topic and don't want to be proved wrong -->

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of QSY facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relyingon morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at QSY, our quantitative analyses clearly differentiate polyhedrons from spheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at QSY represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at QSY, indicating a diversity of reduction goals.

However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores. In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns. At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable. However, it remains unclear whether this typological transformation was intentional or unplanned. In addition, the spheroids at QSY appear to have undergone another form of typological transformation—from hammerstones to spheroids. One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.

@@

In China, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids_distribution_China. In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene. In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai]. The widespread occurrence of spheroids across time and space may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China. However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities. Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China. In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China. Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun]. For example, at the DC site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz. In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].

Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]). Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids]. Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan]. No clear relationship has yet been established between raw material types and specific production methods.

At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun]. MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984]. Of the six sites exhibiting similar strategies to QSY spehroids, all but MJQ are located within the Fen River and Lishui River regions. These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe]. The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy. This is a significant observation, as it implies that spheroid production techniques similar to those at QSY are more frequently associated with lithic assemblages characterized by intentional shaping strategies.

From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites. However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids_distribution_China). Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology. Rather, what we infer is that faceted spheroids in China represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another. If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, this would suggest that shaping techniques originated in an even earlier technological context.

This pattern aligns with spheroid findings in other regions of the world. Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan. Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence]. The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019]. Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020]. In particular, faceted spheroids, which reflect a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
